Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2024-08-18 manuscript/manuscript.Rmd
manuscript update.

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf. Long Core Tools [Shea, -@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was “at least partly under genetic control” [@corbey2016: 6].
Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of function-driven resharpening [@iovita2011; @mcpherron2000] whereas others find it more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
Such debates about the shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions, including the origins of human culture [@corbey2016; @liu2023; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean (ca. 1.7 –1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; Semaw, 2000; Semaw et al., 2003; Stout et al., 2010).
By comparing variation in overall artifact shape (or “Bauplan” [Lycett and Gowlett, 2008)) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.

@@

There is a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
There is much less agreement regarding the less heavily worked and formally standardized LCTs typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018].
Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.

>>>>

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf. Long Core Tools [@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was “at least partly under genetic control” [@corbey2016: 6].
Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of function-driven resharpening [@iovita2011; @mcpherron2000] whereas others find it more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
Such debates about the shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions, including the origins of human culture [@corbey2016; @liu2023; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean (ca. 1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; @semaw2000; @semaw2003; @stoutTechnologicalVariationEarliest2010).
By comparing variation in overall artifact shape (or “Bauplan” [@lycettQuestionsSurroundingAcheulean2008]) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.

@@

There is a broad consensus that refined handaxes and cleavers known from some later Acheulean (\<1.0 Ma) contexts resulted from procedurally elaborate, skill intensive and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014], although debate over the presence of explicit, culturally transmitted shape preferences not determined by function continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
However, archaeologists have also noted the presence of less heavily worked and morphologically standardized LCTs (e.g., handaxes, cleavers, picks) in earlier Acheulean contexts [@leakey1951; @kleindienst1962; @wynn1979; @gowlett1986; @clark1994; @ludwig1998; @stoutStoneToolmakingEvolution2011; @presnyakova2018].
Such forms are typical of Acheulean assemblages in Africa prior to 1.0 Ma [e.g., @kleindienst1962; @ludwig1998; @lepre2011; @beyene2013; @diez-martín2015; @delatorre2018; @presnyakova2018; @semaw2018], but continue to occur with variable frequency in later time periods [@mcnabb2015] and may be especially prevalent in eastern Asian [@li2021].
Although formal tool types are commonly used to describe such early Acheulean assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of reduction.
